---
title: "Class18 Whooping Cough"
author: "Xinyu (A17115443)"
format: pdf
toc: TRUE
---

Pertussis (aka. Whooping cough) is a deadlily lung infection caused by the bacteria B. Pertussis.

The CDC tracks Pertussis cases around the US.
https://www.cdc.gov/pertussis/php/surveillance/pertussis-cases-by-year.html

We can "scrape" this data using the R **datapasta()** package.

```{r, include=FALSE}
cdc <- data.frame(
  year = c(1922L,1923L,1924L,1925L,
           1926L,1927L,1928L,1929L,1930L,1931L,
           1932L,1933L,1934L,1935L,1936L,
           1937L,1938L,1939L,1940L,1941L,1942L,
           1943L,1944L,1945L,1946L,1947L,
           1948L,1949L,1950L,1951L,1952L,
           1953L,1954L,1955L,1956L,1957L,1958L,
           1959L,1960L,1961L,1962L,1963L,
           1964L,1965L,1966L,1967L,1968L,1969L,
           1970L,1971L,1972L,1973L,1974L,
           1975L,1976L,1977L,1978L,1979L,1980L,
           1981L,1982L,1983L,1984L,1985L,
           1986L,1987L,1988L,1989L,1990L,
           1991L,1992L,1993L,1994L,1995L,1996L,
           1997L,1998L,1999L,2000L,2001L,
           2002L,2003L,2004L,2005L,2006L,2007L,
           2008L,2009L,2010L,2011L,2012L,
           2013L,2014L,2015L,2016L,2017L,2018L,
           2019L,2020L,2021L,2022L,2024L),
  cases = c(107473,164191,165418,152003,
                                   202210,181411,161799,197371,
                                   166914,172559,215343,179135,265269,
                                   180518,147237,214652,227319,103188,
                                   183866,222202,191383,191890,109873,
                                   133792,109860,156517,74715,69479,
                                   120718,68687,45030,37129,60886,
                                   62786,31732,28295,32148,40005,
                                   14809,11468,17749,17135,13005,6799,
                                   7717,9718,4810,3285,4249,3036,
                                   3287,1759,2402,1738,1010,2177,2063,
                                   1623,1730,1248,1895,2463,2276,
                                   3589,4195,2823,3450,4157,4570,
                                   2719,4083,6586,4617,5137,7796,6564,
                                   7405,7298,7867,7580,9771,11647,
                                   25827,25616,15632,10454,13278,
                                   16858,27550,18719,48277,28639,32971,
                                   20762,17972,18975,15609,18617,
                                   6124,2116,3044,35493)
)
```

```{r}
head(cdc)
```

```{r}
library(ggplot2)
```

> Q1, 2

```{r}
ggplot(cdc) +
       aes(x= year, y=cases) +
  geom_line() +
  geom_vline(xintercept = 1946, color = "blue") + #introduction of the wP vaccine
  geom_vline(xintercept = 1996, color = "red") + #switch to aP vaccine
  geom_vline(xintercept = 2003, color = "darkgreen") +
  geom_vline(xintercept = 2020, color = "grey") #COVID
```
There were high case numbers before the first wP (whole-cell) vaccine roll out in 1946. Then a rapid decline in case numbers until 2003 when we had our first large-scale outbreaks of pertussis again. Around 2020, case numbers dropped due to COVID masking and lockdown. Then in 2024 there is another increase.

> Q. What is different about the immune response to the infection if you had older wP vaccine vs newer aP vaccine?

## Computational Model of Immunity Pertussis Boost (CMI-PB)

The CMI-PB project aims to address the key question: What's the difference between wP and aP individuals?

We can get all the data from this ongoing project via JSON API calls.
For this we will use the **jsonlite** package.

```{r}
library(jsonlite)
subject <- read_json("https://www.cmi-pb.org/api/v5_1/subject",
                     simplifyVector = TRUE)
head(subject)
```

> Q. How many individuals "subjects" are in this dataset?

```{r}
nrow(subject)
```

> Q4. How many wP and aP primmed indiviuals are in this dataset?

```{r}
table(subject$infancy_vac)
```

> Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```

Obtain more data from CMI-PB:

```{r}
specimen <- read_json("http://cmi-pb.org/api/v5_1/specimen",
                      simplifyVector = TRUE)
ab_data <- read_json("http://cmi-pb.org/api/v5_1/plasma_ab_titer",
                     simplifyVector = TRUE)
```

```{r}
head(subject)
```

```{r}
head(ab_data)
```

I now have 3 tables of data from CMI-PB: `subject`, `specimen`, and `ab_data`.
In order to better examine the data, we use the "*_join()" function from dplyr to combine the tables.
We will use "inner_join()" rather than "full_join()":
"inner_join()" can only join 2 tables at the same time.

> Q9

```{r}
library(dplyr)

meta <- inner_join(subject, specimen)
head(meta)
```

```{r}
dim(subject)
dim(specimen)
```

```{r}
dim(meta)
```

> Q10. Join `meta` and `ab_data` tables:

```{r}
abdata <- inner_join(meta, ab_data)
head(abdata)
```


> Q. How many different isotypes of ab are there in this dataset?

```{r}
length(abdata$isotype)
```

```{r}
table(abdata$isotype)
```

```{r}
table(abdata$antigen)
```

I want a plot of antigen levels across the whole dataset.
MFI:
```{r}
ggplot(abdata) +
  aes(MFI, antigen) +
  geom_boxplot()
```
MFI_normalized:
```{r}
ggplot(abdata) +
  aes(MFI_normalised, antigen) +
  geom_boxplot()
```

Antigens like FIM2/3, PT, and FELD1 have quite a wide range of values. Measles is a control. We expect to see low levels of measles response with a pertussis vaccine. Others like measles don't show much activity.

> Q. Are there differences at this whole-dataset level between aP and wP?

```{r}
ggplot(abdata) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot()
```

```{r}
ggplot(abdata) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot() +
  facet_wrap(~infancy_vac)
```


# 4. Examine IgG Ab titer levels

For this I need to sleect out just isotype IgG.
```{r}
igg <- abdata |>
  filter(isotype == "IgG")
head(igg)
```

A overview boxplot of just the IgG data:

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot()
```

Digging in further to look at the time course of IgG isotype PT antigen levels across aP and wP individuals:

```{r}
## Filter to include 2021 data only
abdata.21 <- abdata |>
  filter(dataset == "2021_dataset")

## Filter to look at IgG PT data only
abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  
## Plot and color by infancy_vac (wP vs aP)
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```









