---
title: "Class10.2Lab"
author: "Xinyu Wen (A17115443)"
format: html
toc: TRUE
---

# 4. Comparative structure analysis of Adenylate Kinase

## Step 1: Search and retrieve ADK structures
```{r}
library(bio3d)
```

Retrive a sequence for the protein we are intersted in. We will take ADK "1ake_A"
The codes does the BLAST search for us, instead of doing it manually.
```{r}
id <- "1ake_A"
aa <- get.seq(id)
aa
```

Run a BLAST search of the PDB for all related sequences to our input `aa`.
```{r}
blast <- blast.pdb(aa)
```

```{r}
hits <- plot(blast)
```

These are out "top hits". i.e. all the structures in the PDB database related to our input sequence.
```{r}
hits$pdb.id
```

## Step 3: 
```{r}
# Download releated PDB files
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)
```


## Step 4: Align and superpose structures

Downloaded BiocManager and msa in Console

```{r}
# Align releated PDBs
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")
```

```{r}
pdbs
```

## Step 5: PCA
Let's use our old friend PCA to make sense o these confussing, complicated structure relationships.

```{r}
# Perform PCA
pc <- pca(pdbs)
```


```{r}
plot(pc)
```

Let's make a trajectory (or movie) of the main conformational changes captured by PC1. We will use the `mktrj()` function for this.
```{r}
mktrj(pc, file="pca_results.pdb")
```
Open this file in molstar viewer.

Back of the envelope comparasion of the PDB size to UniProt
```{r}
pdb <- 231029
uniprot <- 253206171

pdb/uniprot * 100
```








