---
title: "Class 9: Halloween Candy Mini-Project"
author: "Xinyu Wen (A17115443)"
format: pdf
toc: TRUE
---
Today we will examine a data from 538 Halloween candy, using ggplot, dplyr and PCA to understand the multivariate dataset.

# 1. Importing candy data

```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```
> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

> Q2.1. How many chocolate candy?

```{r}
sum(candy$chocolate)
```


# 2. What is your favorate candy?

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Almond Joy", ]$winpercent
```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", "winpercent"]
```

To get a quick overview of a new dataset, we can use the package `skimr`.

```{r}
library("skimr")
skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Data in winpercent row is much higher than other rows.
**N.B.** It looks like the `winpercent` row is on a different scale than the others (0-100% rather than 0-1).

- This row is on a different scale. If we run a PCA, this row will significantly affect the result.


> Q7. What do you think a zero and one represent for the candy$chocolate column?

1 means that this type of candy is a chocolate candy. 0 means this candy is not a chocolate candy.

> Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)
ggplot(candy, aes(winpercent)) +
  theme_replace() +
  geom_histogram(bins = 10, fill = "pink")
```

> Q9. Is the distribution of winpercent values symmetrical?

No it is not symmetrical.

> Q10. Is the center of the distribution above or below 50%?

```{r}
summary(candy$winpercent)
```

The median is 47.83, which is less than 50%


> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

- step 1: Find all "chocolate" candy
- step 2: Find their winpercent value
- step 3: Summarize the value

- step 4: Find all "fruity" candy
- step 5: Find their winpercent value
- step 6: Summarize the value

- step 7: Compare the two summary values.

```{r}
choc.inds <- candy$chocolate == 1 #step 1
choc.win <- candy[choc.inds, ]$winpercent #step 2
choc.mean <- mean(choc.win) #step 3
choc.mean
```

```{r}
fru.inds <- candy$fruity == 1 #step 4
fru.win <- candy[fru.inds, ]$winpercent #step 5
fru.mean <- mean(fru.win) #step 6
fru.mean
```

step 7: Chocolate candy higher ranked/ has higher winpercent value than fruit candy.

> Q12. Is this difference statistically significant?

```{r}
t.test(fru.win, choc.win)
```

Because the p-value is very low, the difference between chocolate and fruity candy is statically significant.



# 3. Overall Candy Rankings

> Q13. What are the five least liked candy types in this set?

```{r}
#Note this is not as useful. It just sort values
sort(candy$winpercent)
```

```{r}
#ex. sort
x <- c(10, 1, 100)
sort(x)
```

```{r}
#ex. order
order(x)
```

```{r}
x[order(x)]
```

The `order()` function tells us how ti arrange the elements of the input to make them sorted - i.e. how to order them.

We can determine the order of winpercent to make them sorted and use that order to arange the whole dataset.

```{r}
ord.inds <- order(candy$winpercent)
head(candy[ord.inds,])
```

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
ord.inds <- order(candy$winpercent, decreasing = T)
head(candy[ord.inds,])
```

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col()
```

Add some useful color to separate types of candies. Make our own color vectors, spell out exactly what candy is represented by which color.

```{r}
mycols <- rep("grey", nrow(candy))
mycols[candy$chocolate ==1] <- "chocolate"
mycols[candy$fruity ==1] <- "green"
mycols[candy$bar ==1] <- "blue"
mycols
```

```{r}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=mycols)
```

> Q17. What is the worst ranked chocolate candy?

Sixlets

> Q18. What is the best ranked fruity candy?

Starburst

# 4. Taking a look at pricepercent

Make a plot of winpercent (x-axis) vs pricepercent (y-axis).

```{r}
ggplot(candy)+
  aes(winpercent, pricepercent,  label=rownames(candy)) +
  geom_point(col=mycols) +
  geom_text()
#Too messy
```

To avoid overplotting, use a package **ggrepel**.

```{r}
library(ggrepel)

ggplot(candy)+
  aes(winpercent, pricepercent,  label=rownames(candy)) +
  geom_point(col=mycols) +
  geom_text_repel(max.overlaps = 6) +
  theme_bw()
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's Miniatures

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
price.ord <- order(candy$pricepercent, decreasing = T)
head(candy[price.ord,])
```

Least popular is Nik L Nip.


# 5. Exploring the correlation structure

First we will use correlation to view
**corrplot** package to plot a correlation.

```{r}
cij <- cor(candy)
cij
```
```{r}
library("corrplot")

corrplot(cij)
```


> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

fruity and chocolate

> Q23. Similarly, what two variables are most positively correlated?

Chocolate and winpercent

# 6. Principal Component Analysis

Let's apply PCA using the `prcomp()` function to our candy dataset remembering to set the **scale=TURE**

```{r}
pca <- prcomp(candy, scale = TRUE)
```

```{r}
summary(pca)
```

```{r}
attributes(pca)
```

Let's plot the main results as our PCA "score plot"

```{r}
ggplot(pca$x) +
  aes(PC1, PC2, label = rownames(pca$x)) +
  geom_point(col=mycols) +
  geom_text_repel(col=mycols, max.overlaps = 12)
```



Finally let's look at how the origional variable s contribute tot he PCs, start with PC1

```{r}
ggplot(pca$rotation, aes(PC1, reorder(rownames(pca$rotation), PC1)))+
  geom_col()
```


> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Bars towards the right means they are more "fruity". Bars towards the left means they are more "chocolate" and "bar". They also have higher "winpercent" and "pricepercent".




